New Bayes covariant coupling methods to cope with change of priors in one-vs-one classification

### 2025-08-06 coupling.tex
Fix multi-class to multiclass

A multi-class estimate $\hat{\boldsymbol{p}}$ obtained by a coupling method implies via \eqref{eq:bt1} probability distributions on each pair of classes:
\begin{align*}
	\boldsymbol{d}_{ij}^\textrm{multi}= \biggl(\frac{\hat{p_i}}{\hat p_i + \hat p_j},\frac{\hat{p_j}}{\hat p_i + \hat p_j}\biggr)	
\end{align*}
On the other hand, binary classifiers provide another distribution for each pair of  classes, namely
\begin{align*}
\boldsymbol{d}_{ij}^\textrm{binary} = ( r_{ij},  r_{ji}).
\end{align*}

@@

Suppose first that $a_1, \ldots,a_M$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_M$ are probabilistic multi-class classification methods satisfying Assumption \ref{ass:2}. We define their linear combination $\bigoplus_i a_m \boldsymbol{v}_m$ as the classification method that for any $x$ yields the probability distribution on $K$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_M \boldsymbol{v}_M\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication, and exponentials are computed component-wise. With this definition we can state the following result.

>>>>

A multiclass estimate $\hat{\boldsymbol{p}}$ obtained by a coupling method implies via \eqref{eq:bt1} probability distributions on each pair of classes:
\begin{align*}
	\boldsymbol{d}_{ij}^\textrm{multi}= \biggl(\frac{\hat{p_i}}{\hat p_i + \hat p_j},\frac{\hat{p_j}}{\hat p_i + \hat p_j}\biggr)	
\end{align*}
On the other hand, binary classifiers provide another distribution for each pair of  classes, namely
\begin{align*}
\boldsymbol{d}_{ij}^\textrm{binary} = ( r_{ij},  r_{ji}).
\end{align*}

@@

Suppose first that $a_1, \ldots,a_M$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_M$ are probabilistic multiclass classification methods satisfying Assumption \ref{ass:2}. We define their linear combination $\bigoplus_i a_m \boldsymbol{v}_m$ as the classification method that for any $x$ yields the probability distribution on $K$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_M \boldsymbol{v}_M\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_M^{a_M}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication, and exponentials are computed component-wise. With this definition we can state the following result.

### 2025-08-06 experiment.tex
Fix multi-class to multiclass

The choice of a coupling method is an important task for practitioners. We aim to help by posing the following interrelated questions

@@

These questions will be put to test in Sections \ref{sec:exp1} --\ref{sec:exp4}. The methodology of our experiments is describe in the next two subsections.

@@

It is natural to consider the benchmark datasets used in previous evaluation of coupling methods [\cite{wu2004probability}]. The authors considered altogether seven datasets with number of classes ranging from 3 to 26. For each dataset they extracted 40 pairs of training and testing datasets. Half of these were smaller (300 training samples and 500 testing samples) and the other half was larger (800 training samples and 1000 testing samples). In our experiments we use only the larger pairs.

>>>>

The choice of a coupling method is an important decision for practitioners. We aim to investigate the following interrelated questions

@@

These questions will be put to test in Sections \ref{sec:exp1} --\ref{sec:exp4}. The methodology of our experiments is described in the next two subsections.

@@

It is natural to consider the benchmark datasets used in previous evaluation of coupling methods \cite{wu2004probability}. The authors considered altogether seven datasets with number of classes ranging from 3 to 26. For each dataset they extracted 40 pairs of training and testing datasets. Half of these were smaller (300 training samples and 500 testing samples) and the other half was larger (800 training samples and 1000 testing samples). In our experiments we use only the larger pairs.

### 2025-08-06 theory.tex
Fix multi-class to multiclass

Suppose we have $K$ classes of objects, each distributed with density $d_i$, $i=1,\ldots, K$ on a feature 
space $X$. In a multi-class classification problem we are provided with 
the prior probabilities $\Pi = (P_1, \ldots, P_K)$ of each class. Given $x$ in $X$ we aim to find 
the probability that $x$ belongs to the $i$-th class.

@@

Besides the multi-class classifier we may also consider the binary Bayes classifiers for any pair $(i,j)$ of 
classes. Assuming the prior on the two classes is proportional to $(P_i, P_j)$, the binary Bayes classifier 
predicts the probability of the $i$-th class as

@@

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{observation $x$}\arrow[r, "binary~classifiers"] \arrow[dr,dotted,swap, "multi-class~ prediction"] & \boldsymbol{R}(x) %\arrow{d}{coupling}[swap]{\boldsymbol{v}} \\
		\arrow{d}{coupling} \\
		 & \boldsymbol{v}(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multi-class classification using a coupling method $\boldsymbol{v}$}
	\label{fig:coupling}
\end{figure}

@@

\emph{Unique solution} refers to the fact that the procedure to compute a multi-class probabilistic prediction always converges to a unique solution. One way to satisfy this condition would be that it amounts to a linear system of equations in $K$ variables, and the matrix of the system would be invertible. Another example would be when the procedure amounts to finding the minimum of a strongly convex function.

\emph{Bradley-Terry consistency} refers to the outcome of a procedure providing a unique solution. The requirement states that should we apply the coupling procedure to binary Bayes classifiers, then the result is the output of multi-class Bayes classifier i.e. satisfies \eqref{eq:bt1}.

@@

\emph{Symmetry} refers to transformations of multi-class predictions when permuting classes. We can define two natural actions on permutation group on $K$-elements. Suppose $\sigma$ is a permutation on $K$ elements.

@@

\emph{Bayes covariance} is a notion introduced in  \cite{vsuch2016bayes}. It refers to the behavior of a probabilistic multi-class method built in one-vs-one fashion, when the priors change. Any Bayes classifier, whether binary or multiclass changes its prediction when the priors change. Let us describe the process. Suppose we change priors from $\Pi = (P_1, \ldots, P_k)$ to $\Pi'= (P'_1, \ldots, P'_K)$. Then the prediction of Bayes classifiers changes from $\boldsymbol{p}^\Pi= (p_1, \ldots, p_K)$ to the vector 
\begin{align}
\boldsymbol{p}^{\Pi'} \propto \biggl(p_1 \frac{P'_1}{P_1}, \ldots, p_K \frac{P'_K}{P_K}\biggr). \label{eq:changePrior}
\end{align}

@@

Let us finally mention \emph{Hinton's minimality}. Geoffrey Hinton is reported to object to the underlying principle of one-vs-one classification, where any single classifier influences the multi-class prediction \cite[p.~467]{hastie1998classification}. Suppose we are classifying objects from CIFAR-10 dataset.  If the true class is  a ``dog'', why should we care what is the output of the classifier distinguishing ``airplanes'' from ``ships''? The binary classifier is trained only on airplanes and ships, and has never seen a dog. One may thus prefer a classifier which somehow suppresses noisy output of (seemingly) irrelevant classifiers.

>>>>

Suppose we have $K$ classes of objects, each distributed with density $d_i$, $i=1,\ldots, K$ on a feature 
space $X$. In a  classification problem we are provided with 
the prior probabilities $\Pi = (P_1, \ldots, P_K)$ of each class. Given $x$ in $X$ we aim to find 
the probability that $x$ belongs to the $i$-th class.

@@

Besides the multiclass classifier we may also consider the binary Bayes classifiers for any pair $(i,j)$ of 
classes. Assuming the prior on the two classes is proportional to $(P_i, P_j)$, the binary Bayes classifier 
predicts the probability of the $i$-th class as

@@

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{observation $x$}\arrow[r, "binary~classifiers"] \arrow[dr,dotted,swap, "multiclass~ prediction"] & \boldsymbol{R}(x) %\arrow{d}{coupling}[swap]{\boldsymbol{v}} \\
		\arrow{d}{coupling} \\
		 & \boldsymbol{v}(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multiclass classification using a coupling method $\boldsymbol{v}$}
	\label{fig:coupling}
\end{figure}

@@

\emph{Unique solution} refers to the fact that the procedure to compute a multiclass probabilistic prediction always converges to a unique solution. One way to satisfy this condition would be that it amounts to a linear system of equations in $K$ variables, and the matrix of the system would be invertible. Another example would be when the procedure amounts to finding the minimum of a strongly convex function.

\emph{Bradley-Terry consistency} refers to the outcome of a procedure providing a unique solution. The requirement states that should we apply the coupling procedure to binary Bayes classifiers, then the result is the output of multiclass Bayes classifier i.e. satisfies \eqref{eq:bt1}.

@@

\emph{Symmetry} refers to transformations of multiclass predictions when permuting classes. We can define two natural actions on permutation group on $K$-elements. Suppose $\sigma$ is a permutation on $K$ elements.

@@

\emph{Bayes covariance} is a notion introduced in  \cite{vsuch2016bayes}. It refers to the behavior of a probabilistic multiclass method built in one-vs-one fashion, when the priors change. Any Bayes classifier, whether binary or multiclass changes its prediction when the priors change. Let us describe the process. Suppose we change priors from $\Pi = (P_1, \ldots, P_k)$ to $\Pi'= (P'_1, \ldots, P'_K)$. Then the prediction of Bayes classifiers changes from $\boldsymbol{p}^\Pi= (p_1, \ldots, p_K)$ to the vector 
\begin{align}
\boldsymbol{p}^{\Pi'} \propto \biggl(p_1 \frac{P'_1}{P_1}, \ldots, p_K \frac{P'_K}{P_K}\biggr). \label{eq:changePrior}
\end{align}

@@

Let us finally mention \emph{Hinton's minimality}. Geoffrey Hinton is reported to object to the underlying principle of one-vs-one classification, where any single classifier influences the multiclass prediction \cite[p.~467]{hastie1998classification}. Suppose we are classifying objects from CIFAR-10 dataset.  If the true class is  a ``dog'', why should we care what is the output of the classifier distinguishing ``airplanes'' from ``ships''? The binary classifier is trained only on airplanes and ships, and has never seen a dog. One may thus prefer a classifier which somehow suppresses noisy output of (seemingly) irrelevant classifiers.
